---
title: "Effect of Potassium Fertilizer Rates on Soybean Trifoliate Concentration at Four different Location"
author: "Faith Oyewale OLABISI"
format: pdf
editor: visual
---

## Introduction

This study was conducted at four different locations in one year to assess the effect of potassium (K) fertilizer rates on soybean trifoliate K concentration at R4.

Each study was a randomized complete block design (RCBD) with K rate as the only treatment factor with levels 0, 50, 100, and 150 lbs K20/ac.

Note: The experiment work analysed in this project is a simulated experiment.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(car)
library(knitr)
library(readxl)
library(openxlsx)
library(lme4)
library(broom.mixed)
library(ggthemes)
library(emmeans)
library(multcomp)
library(multcompView)
```

```{r}
soy_data <- read_excel('data/SoybeanWorkshop.xlsx', sheet = 'R')

view(soy_data)
```

## Data Insight and Exploration

```{r}
soy_data |> 
  glimpse()
```

```{r}
soy_data |> 
  str()
```

```{r}
soy_data |> 
  filter(
    Krate_lbac == 50 | Krate_lbac == 100
  ) 
```

```{r}
soy_data |> 
  head(n = 10) |> 
  kable(
    caption = 'First Ten rows of the data'
  )
  
```

```{r}
soy_data |> 
  summary() |> 
  kable(
    caption = 'Summary of the data'
  )
```

```{r}
soy_data |> 
  summarise(
    mean_A1 = mean(A1.Kpct),
    mean_O2 = mean(O2.Kpct),
    mean_S1 = mean(S1.Kpct)
  ) |> 
  kable(
    caption = 'Mean of the three Plots'
  )
```

```{r}
soy_data |> 
  names()
```

## Data Preparation for Analysis

In this chunk, we wrangled the data to best fit to perform data analysis and randomized complete block design

```{r}
soy_data_final <- soy_data |>
  ## Convert the Krate from lbs/ac to Kg per hectare
  ## Represent the plot between 100 and 400 with a value of 1 to 4
  mutate(
    krate_kgha = Krate_lbac*0.453592/0.4044686,
    krate_kgha = round(krate_kgha, 0),
    Rep = case_when(
      Plot > 100 & Plot < 200 ~ 1,
      Plot > 200 & Plot < 300 ~ 2,
      Plot > 300 & Plot < 400 ~ 3,
      Plot > 400 ~ 4
    )
  ) |>
  ## Gather the location columns into a single column called 'location.k'
  gather(
    Location.k, K_pct, 
    A1.Kpct, O2.Kpct, S1.Kpct, W1.Kpct
  ) |> 
  ## Separate the location.k column into two different column
  ## called location and kname
  separate(
    Location.k,
    into = c('Location', 'Kname')
  ) |> 
  ## select the columns that will be needed to perform the analysis
  dplyr::select(
    Location, Rep, krate_kgha, K_pct
  )
```

```{r}
soy_data_final |> 
  ## Group the data by the location of the site.
  group_by(
    Location
  )|> 
  summarise(
    meanK_pct = mean(K_pct),
  )
```

## Data Visualization

```{r}
## Visualize the rate of potassium application on the various site
soy_data_final |> 
  ggplot(aes(as.factor(krate_kgha), K_pct))+
  geom_boxplot()+
  labs(
     x = "Rate of Potassium {krate_kgha}",
     y = 'K_pct',
     caption = '@FaithOyewaleOLABISI'
  )+
  theme_minimal()
```

```{r}
## Visualize the rate of potassium application on the various site or location
soy_data_final |> 
  ggplot(aes(as.factor(krate_kgha), K_pct))+
  geom_boxplot(aes(fill = Location))+
  theme_minimal()
```

```{r}
soy_data_final |> 
  ggplot(aes(as.factor(krate_kgha), K_pct))+
  geom_boxplot(aes(fill = Location))+
  facet_grid(.~Location)+
  theme_minimal()
```

```{r}
soy_data_final |> 
  ggplot(aes(K_pct))+
  geom_density(color = 'red')+
  theme_minimal()
```

```{r}
# Creating krate as categorical variable(factor) and other columns 
# also to factor
soy_data_final_1 <- soy_data_final |> 
  mutate(
    fkrate_kgha = factor(krate_kgha),
    frep = factor(Rep),
    Location = factor(Location)
  ) |> 
  filter(
    Location == 'W1' | frep != '1'
  )
```

## Randomized Complete Block Design (RCBD) Model

```{r}
# Changing type of matrix restriction
options(contrasts = c('contr.sum', 'contr.poly'))

# Running the Model
soyk_mod <- lmer(K_pct ~ Location*fkrate_kgha + (1|Location/frep),
                 data = soy_data_final_1)
soyk_mod
```

```{r}
Anova(soyk_mod, type = 3)
```

## Model Assumption

```{r}
## Extracting residual information

soy_data_aug <- augment(soyk_mod)

## Adding pearson standardized residuals

soy_data_aug$.stdresid <- resid(soyk_mod, type = 'pearson',
                                scale = 1)
```

### Within-group errors are iid \~ N(0, var2)

On this plot, looking for:\
- Spread around zero\
- Homogeneity (no patterns)\
- Outliers (\>3 or \<-3)

```{r}
soy_data_aug |> 
  ggplot(aes(.fitted, .stdresid))+
  geom_point(aes(color = Location))+
  geom_hline(yintercept = c(-3, 0, 3), color = 'orange')+
  geom_smooth()+
  theme_minimal()
```

On this plot, looking for normality (points on top of line).

```{r}
soy_data_aug |> 
  ggplot(aes(sample = .stdresid))+
  stat_qq()+
  stat_qq_line()+
  theme_minimal()
```

### Random effects are iid \~ N(0,var1)

On this plot, looking for normality.

```{r}
randeff_LocRep <- ranef(soyk_mod)[[1]]

randeff_LocRep |> 
  ggplot(aes(sample = '(intercept)'))+
  stat_qq()+
  stat_qq_line()+
  theme_minimal()

```

```{r}
randeff_rep <- ranef(soyk_mod)[[2]]

randeff_rep |> 
  ggplot(aes(sample = '(intercept)'))+
  stat_qq()+
  stat_qq_line()+
  theme_minimal()
```

### Extracting model means and pairwise comparisons

```{r}
soyk_means_LocRate <- emmeans(soyk_mod, 
                              ~ fkrate_kgha | Location)

soyk_means_LocRate
```

```{r}
soyk_pwc_LocRate<- cld(soyk_means_LocRate, 
                       adjust="none",
                       Letters=letters,
                       reversed=T)

soyk_pwc_LocRate <- soyk_pwc_LocRate |> 
  as.data.frame()

soyk_pwc_LocRate
```

```{r}
soyk_pwc_LocRate |> 
  ggplot(aes(fkrate_kgha, emmean))+
  geom_bar(aes(fill = Location), color = 'black', 
           stat = 'identity', alpha = 0.8)+
  geom_text(aes(label = .group), nudge_y = 0.3, 
            show.legend = F)+
  labs(
    x = 'K Rate (kg k2o/ha)',
    y = 'Plant K (%)',
    caption = '@Faith Oyewale OLABISI'
  )+
  theme_minimal()+
  scale_fill_tableau()+
  theme(legend.position = 'none')+
  scale_y_continuous(limits = c(0, 4))+
  facet_wrap(~Location)
```

```{r}
soy_data_final_1 |> 
  ggplot(aes(fkrate_kgha, K_pct))+
  geom_boxplot(aes(fill = Location), color = 'black', alpha = .8)+
  facet_wrap(~Location, scales = 'free')+
  scale_y_continuous(limits = c(.5, 4))+
  labs(
    x = 'K Rate (kg K2O/ha)',
    y = 'Plant K (%)',
    caption = '@FaithOyewaleOLABISI'
  )+
  theme_minimal()+
  scale_fill_tableau()+
  theme(legend.position = 'none')


```

## Acknowledgement

Kansas State University 2019 AGSA R Workshop - Introduction to R, taught by [\@leombastos](https://github.com/leombastos).
