---
title: "Estimation of Rainbow Trout Abundance in Upper Niagra Springs Pond using the Single Census Mark-Recapture Methods"
author: "Faith Oyewale OLABISI"
format: pdf
editor: visual
toc: true
toc-depth: 5
---

## Introduction

Warren *et al*. (2004) examined the population of Rainbow Trout (Oncorhynchus mykiss) in the Upper Niagara Springs Pond in 2000. Fish were captured at two times by using an electrofishing unit attached to a driftboat. The capture history of all fish examined in the two samples that were 100 mm and longer is in RBTroutUNSP.

1.  Use these data to answer the following questions.

-   Create a summary of the capture histories. From your capture history summary assign values to each of M, n, and m Construct an appropriate population estimate, with a 95% confidence interval, for Upper Niagara Springs Pond Rainbow Trout in 2000.

-   Carefully interpret the results. Which method did you use to construct the confidence interval? Explain why you chose that method.

## Data Collection and Importation

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(car)
library(knitr)
```

```{r}
#| message: false
#| warning: false
trout_data <- read_csv('data-raw/RBTroutUNSP.csv')
view(trout_data)
trout_data |> 
  str()
```

## Summary of Capture of Histories

Summary of the Rainbow Trout Abundance data

```{r}
trout_data |> 
  summary() |> 
  kable(
    align = 'lccr',
    caption = 'Summary of the Rainbow Trout Abundance'
  )
```

Summary of the First Time Capture History

```{r}
trout_data |> 
  summarise(
    mean_first_year = mean(first),
    median_first_year = median(first),
    max_first_year = max(first),
    min_first_year = min(first)
  ) |> 
  kable(
    caption = 'Summary of the First Time Capture History'
  )
```

Summary of the Second Time Capture History

```{r}
trout_data |> 
  summarise(
    mean_second_year = mean(second),
    median_second_year = median(second),
    max_second_year = max(second),
    min_second_year = min(second)
  ) |> 
  kable(
    caption = 'Summary of the Second Time Capture History'
  )
```

Number of Tagged and Un-tagged Fish caught the First time

```{r}
trout_data |> 
  group_by(first) |> 
  summarise(count = n()) |> 
  kable(
    caption = 'Summary of Tagged and Untagged Fish caught the First time'
  )
```

Number of Tagged and Un-tagged Fish caught the Second time

```{r}
trout_data |> 
  group_by(second) |> 
  summarise(count = n ()) |> 
  kable(
    caption = 'Summary of Tagged and Untagged Fish caught the Second Time'
  )
```

```         
```

## Construction of the Population Estimate Key using the Lincoln-Peterson Estimator

In this study:

-   N is the total number of fish in the population.

-   C is the total number of fish captured in both samples.

-   R is the number of recaptures.

To assign these values, we can calculate them directly from our data:

```{r}
# Assign values to N, C, and R
C <- nrow(trout_data) # Total captures from both samples
R <- sum(trout_data$first == 1 & trout_data$second == 1) # Sum of Recaptures
N <- C + R # Total population estimate (using lincoln-petersen estimator)
```

## Calculation of Population Estimate using Lincoln-Petersen Estimator

Using the Lincoln-Petersen estimator, we can estimate the population size and calculate a confidence interval. The formula for estimating population size is:

N = (C~1~ + 1) (C~2~ + 1) / (R + 1) - 1

Where:

-   C​ is captures in the first sample,

-   C​ is captures in the second sample,

-   R is recaptures.

We will also calculate a confidence interval using a normal approximation method.

```{r}
# Calculate population estimate using Lincoln-Petersen Estimator
C1 <- sum(trout_data$first == 1)
C2 <- sum(trout_data$second == 1)

estimate_size <- ((C1 + 1) * (C2 + 1)) / (R + 1) - 1
```

## Calculation of Standard Error and Confidence Interval

```{r}
# calculate standard error for confidence interval
SE <- sqrt((C1 * C2 * (C1 - R) * (C2 - R)) / ((R + 1)^2 * (R + 2)))

# Calculate Confidence Interval
con_lower <- estimate_size - qnorm(0.975) * SE
con_upper <- estimate_size + qnorm(0.975) * SE
```

## Results

```{r}
estimate_size
con_lower
con_upper 
```

## Discussion

The normal approximation method was chosen for constructing the confidence interval because it is suitable when sample sizes are sufficiently large to meet normality assumptions. This method provides a straightforward way to calculate confidence intervals based on standard errors derived from sample data, making it effective for estimating fish populations in ecological studies.

## Conclusion

The estimated population size of Rainbow Trout in Upper Niagara Springs Pond in 2000 is approximately 693. The confidence interval indicates that we are 95% confident that the true population size lies between 370 and 1020​.

## References

1.  Ogle, D. H. (2018). Introductory Fisheries Analyses with R. United States: CRC Press.
2.  Warren, C. D., Frank, K. A. and Partridge. F. E. (2004). [Regional fisheries management investigations - Magic Valley region](https://collaboration.idfg.idaho.gov/FisheriesTechnicalReports/Mgt04-04Warren2001%20Regional%20Fisheries%20Management%20Investigations%20Magic%20Valley%20Region.pdf). Completion Report, Idaho Department of Fish and Game.
